---
title: "Dissolved Oxygen and Aquatic Invertebrate Abundance at North Campus Open Space"
author: "Stella Freedberg, Kimberly Wong, Sami Gibson"
date: today
format: 
  pdf:
    toc: true 
bibliography: references.bib 
execute: 
  echo: false 
  warning: false 
  message: false 
  results: hide ---
---
```{r}
#| label: packages
# load in tidyverse package
library(tidyverse)
# load in here package
library(here)
# load in janitor package
library(janitor)
# load in snakecase package
library(snakecase)
# load in scales package
library(scales)
# load in readxl package
library(readxl)
# load in naniar package
library(naniar)
# load in patchwork package
library(patchwork)
# load in dplyr package
library(dplyr)
# load in broom package
library(broom)
# load in corrplot package
library(corrplot)
# load in ggrepel package
library(ggrepel)
```
```{r}
#| label: read-in-data
# read in taxonomic info data
taxon_list <- read_csv(here("data", "taxon_list.csv"))

# read in aquatic ivert. data
aq_ins <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                    sheet = "Aquatic Insects")

# read in site data
sites <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                   sheet = "sites")

# read in water quality data
water_quality <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                           sheet = "Water Quality",
                           na = c("", "n/a", "N/A", "over", "173+")) 
```
```{r}
#| label: Creating-Consistent-Aesthetics
# shared color palette for taxon-based figures
taxon_colors <- c(
  "Copepod" = "#1B9E77",
  "Ostracod" = "#D95F02",
  "Cladocera" = "#7570B3",
  "Hemiptera Corixidae Boatman" = "#E7298A",
  "Nematode" = "#66A61E")

# shared color palette for DO-category figures
do_colors <- c(
  "low" = 'red3',
  "moderate" = "white",
  "high" = 'blue3')

# shared colors for model lines, summary points, and raw points
model_color <- "#4C78A8"
summary_color <- "#222222"
raw_point_color <- "#333333"
ci_fill <- "#A6CEE3"
reference_color <- "gray45"

# shared theme for all figures
theme_ncos <- theme_classic() +
  theme(
    plot.title = element_text(face = "bold", size = 11),
    axis.title = element_text(size = 8),
    axis.text = element_text(size = 6),
    strip.background = element_rect(fill = "gray90", color = "gray50"),
    strip.text = element_text(face = "bold"),
    legend.position = "right"
  )
```
```{r}
#| label: cleaning-sites-object
# creating new object from sites
ncos_sites <- sites |> 
  # filter to only include NCOS sites sampled four times a year
  filter(sector == "NCOS" & sampling_frequency == "Quarterly")
```
```{r}
#| label: cleaning-taxon-list
# creating new clean object from taxon_list
taxon_list_clean <- taxon_list |> 
  # converting taxon name to snake case (no spaces or capital letters,
  # only underscores)
  mutate(verbatim_name = to_any_case(verbatim_name, case = "snake"))
```
```{r}
#| label: water-quality-cleaning

# creating new clean object from water quality
water_quality_clean <- water_quality |> 
  # clean column names
  clean_names() |> 
  # filtering join: only include sites in ncos_sites
  semi_join(ncos_sites, by = "site") |> 
  # create new column to join with later data frames
  unite("date_site", date, site, remove = TRUE) |> 
  # group by date_site column
  group_by(date_site) |>
  # calculate median pH, salinity, DO
  summarize(med_ph = median(dissolved_oxygen_mg_l, na.rm = TRUE),
            med_sal = median(salinity_ppt, na.rm = TRUE),
            med_do = median(dissolved_oxygen_mg_l, na.rm = TRUE)) |> 
  # ungroup data frame
  ungroup() |> 
  # filter out salinity outliers
  filter(date_site != "2022-08-12_NVBR") 
```
```{r}
#| label: cleaning-aq-ins

# creating new clean object from aquatic inverts
aq_ins_clean <- aq_ins |> 
  # clean column names
  clean_names() |> 
  # filtering join: only include sites occurring in ncos_sites
  semi_join(ncos_sites, by = c("site")) |> 
  # renaming columns
  rename(date = date_on_vial) |> 
  # select columns of interest
  select(date, sample_type, site,
         ostracod,
         copepod,
         hemiptera_corixidae_boatman,
         cladocera,
         nematode,
         diptera_ceratopogonidae, 
         annelida_oligochaete,
         diptera_chironomid,
         annelida_polychaete) |> 
  # filter to only include "filtered beaker" samples
  filter(sample_type %in% c("FB 250", "FB250")) |> 
  # convert the data frame to long format
  pivot_longer(cols = ostracod:annelida_polychaete,
               names_to = "taxon",
               values_to = "abundance") |> 
  # group by date, site, taxon
  group_by(date, site, taxon) |> 
  # calculate average abundance per liter (sum abundance divided by 7.5 liters)
  summarize(ave_lit = sum(abundance, na.rm = TRUE)/7.5) |> 
  # ungroup data frame
  ungroup() |> 
  # create a new column called `date_site` to join with water quality
  unite("date_site", date, site, remove = FALSE) |> 
  # join with water quality
  left_join(water_quality_clean, by = c("date_site")) |> 
  # join with taxon list
  left_join(taxon_list_clean, by = c("taxon" = "verbatim_name")) |> 
  # add full names of sites
  mutate(site_full = case_when(
    site == "NVBR" ~ "North of Venoco Bridge",
    site == "NEC" ~ "South of Venoco Bridge",
    site == "NMC" ~ "Main Channel",
    site == "NPB" ~ "South of Phelps Bridge",
    site == "NPB1" ~ "North of Phelps Bridge",
    site == "NPB2" ~ "Phelps Road",
    site == "NWP" ~  "West Pond",
    site == "NDC" ~ "Devereux Creek"
  )) |> 
  # setting factor levels for site
  mutate(site_full = fct_reorder(site_full, med_do, .fun = "median", .na_rm = TRUE),
         site_full = fct_rev(site_full))
```
```{r}
#| label: creating-focus_taxa
# taxa to include in all figures
focus_taxa <- c("Ostracod", "Copepod", "Cladocera", 
                "Hemiptera Corixidae Boatman", "Nematode")
```
```{r}
#| label: creating-DO_taxa-total_abundance-and-DO-bins
# create a new objecy aq_ins_clean_bins that adds DO bins to aq_ins_clean
aq_ins_clean_bins <- aq_ins_clean |> 
  # create a new bins column based on site_full values
  mutate(bins = case_when(site_full == "Phelps Road" ~ "low",
                             site_full == "North of Phelps Bridge" ~ "low", 
                             site_full == "South of Venoco Bridge" ~ "moderate",
                             site_full == "South of Phelps Bridge" ~ "moderate",
                             site_full =="Devereux Creek" ~ "moderate",
                             site_full =="North of Venoco Bridge" ~ "high",
                             site_full == "Main Channel" ~ "high")) |> 
  # replace underscores in taxon with spaces & convertto title case
  mutate(taxon = str_replace_all(taxon, "_", " "), taxon = str_to_title(taxon)) |>
  # keep only taxa included in the focus_taxa object
  filter(taxon %in% focus_taxa) |>
  # create column that calculates log-transformed abundance
  mutate(log_ave_lit = log(ave_lit + 1)) |> 
  # convert bins into an ordered factor 
  mutate(bins = factor(bins, levels = c("low", "moderate", "high")))

# create object DO_taxa from aq_ins_clean
DO_taxa <- aq_ins_clean |> 
  # remove rows that are missing median salinity values
  filter(!is.na(med_do)) |> 
  # create column that calculates log-transformed abundance
  mutate(log_ave_lit = log(ave_lit + 1)) |>
  # mutates to title-case taxon names
  mutate(taxon = str_replace_all(taxon, "_", " "), taxon = str_to_title(taxon)) |>
  # only include focus taxa
  filter(taxon %in% focus_taxa) |> 
  # order taxon levels by maximum average abundance per liter
  mutate(taxon = fct_reorder(taxon, ave_lit, .fun = max, .desc = TRUE)) |> 
  # keep only the columns needed for the figure
  select(date, site, site_full, taxon, ave_lit, log_ave_lit, med_do)

# create objetc total_abundance from DO_taxa
total_abundance <- DO_taxa |>
  # keep only taxa included in the focus_taxa object
  filter(taxon %in% focus_taxa) |>
  # group observations by date, site, and median DO
  group_by(date, site, med_do) |>
  # sum average abundance per liter across taxa & ignore missing values
  summarize(total_abundance = sum(ave_lit, na.rm = TRUE),
            # ungroup
            .groups = "drop") |>
  # create a log10-transformed total abundance column called log_total_abundance
  mutate(log_total_abundance = log10(total_abundance + 1))

# updates the total_abundance object
total_abundance <- total_abundance |>
  # create a new variable do_category based on DO quantiles
  mutate(
    # classify each observation into low, medium, or high dissolved oxygen
    do_category = case_when(
      # assign Low DO to values at or below the 25th percentile of median DO
      med_do <= quantile(med_do, 0.25, na.rm = TRUE) ~ "Low DO",
      # assign Medium DO to values between the 25th percentile - 75th percentile
      med_do <= quantile(med_do, 0.75, na.rm = TRUE) ~ "Medium DO",
      # assign High DO to all remaining values above the 75th percentile
      TRUE ~ "High DO"))

# edit the column do_category into an ordered factor
total_abundance$do_category <- factor(total_abundance$do_category,
                                      # set the category order
                                      levels = c("Low DO", "Medium DO", "High DO"))
```
# Introduction

# Methods

# Results

We first examined dissolved oxygen patterns across NCOS sampling sites to understand the oxygen conditions experienced by aquatic invertebrates. Median dissolved oxygen varied substantially among sites, indicating that invertebrates were not sampled under uniform oxygen conditions across the study area (Figure 1). Most median DO values were concentrated around approximately 8–11 mg/L, but the full range extended from very low values to values above 20 mg/L. Phelps Road and North of Phelps Bridge had relatively low median DO values, while Main Channel had the highest median DO values. This site-level variation provides important context for the main research question because it shows that the dataset includes a clear gradient in dissolved oxygen conditions.

## Figure 1. Distribution of Median DO Across Sites

```{r}
#| label: figure-1
# create object aq_ins_clean_fig2 from aq_ins_clean
aq_ins_clean_fig2 <-  aq_ins_clean |>
  # convert taxon to title case
  mutate(taxon = str_replace_all(taxon, "_", " "), taxon = str_to_title(taxon)) |> 
  # filter to only include focus taxa
  filter(taxon %in% focus_taxa)

# create new object median_fig2 from aq_ins_clean_fig2
median_fig2 <- aq_ins_clean_fig2 |> 
  # group by site_full
  group_by(site_full) |> 
  # calculate the median DO value for each site
  summarize(median_do = median(med_do, na.rm = TRUE))

# boundaries of quartiles
q <- quantile(aq_ins_clean_fig2$med_do, probs = seq(0, 1, 0.25), na.rm = T)

# create new column in aq_ins_clean_fig2 that assigns quartiles for each site
aq_ins_clean_fig2$quartile <- cut(
  # use median DO values to assign quartile categories
  aq_ins_clean_fig2$med_do, 
  # define quartile breaks using the 0%, 25%, 50%, 75%, and 100% quantiles
  breaks = quantile(aq_ins_clean_fig2$med_do, probs = seq(0, 1, 0.25),
                    na.rm = TRUE), # discludes NA values
  labels = c("Q1", "Q2", "Q3", "Q4"), # label each quartile group
  include.lowest = TRUE, # include the lowest DO value in the 1st quartile 
  right = FALSE) # make intervals left-closed and right-open

# create new object median_fig2 from aq_ins_clean_fig2
median_fig2 <- aq_ins_clean_fig2 |> 
  # group by site_full
  group_by(site_full) |> 
  # calculate the median DO value for each site
  summarize(median_do = median(med_do, na.rm = TRUE))

# boundaries of quartiles
q <- quantile(aq_ins_clean_fig2$med_do, probs = seq(0, 1, 0.25), na.rm = T)

# create new object label_df from aq_ins_clean_fig2
label_df <- aq_ins_clean_fig2 |> 
  # group the data by site so each site has one label location
  group_by(site_full) |> 
  # calculate the median DO value for each site
  summarize(median_do = median(med_do, na.rm = TRUE))


# create a small data frame for quartile labels
quartile_df <- data.frame(
  x = q, # store the original quartile x-values
  # create adjusted x-positions so quartile labels do not overlap or get cut off
  x_label = c(q[1] + 0.4, q[2] + 0.4, q[3] - 0.3, q[4] + 0.3, q[5] - 0.4),
  # set all quartile labels near the top of the plot
  y = c(0.148, 0.148, 0.148, 0.148, 0.148),
  # create label text showing the quartile name and rounded DO value
  label = paste0(names(q), "\n", round(q, 2)))

# create distribution plot of median dissolved oxygen levels across sites
ggplot(aq_ins_clean_fig2, aes(x = med_do)) +
  # add density curve showing the distribution of median DO values
  geom_density(fill = ci_fill,
               alpha = 0.4, # sets transparency 
               color = model_color, # sets color
               linewidth = 1.2) +  # set density curve line thickness
  # add vertical dashed lines at the quartile values
  geom_vline(xintercept = q,
             linetype = "dashed", # make quartile lines dashed
             color = reference_color) + # set consistent color for quartile lines
  # add short vertical segments marking the median DO value for each site
  geom_segment(data = label_df, # use label_df data
               aes(x = median_do, # map segment positions to site median DO values
                   xend = median_do, # ending x-position at median DO value
                   y = 0, # start each segment at the bottom of the plot
                   yend = 0.012), # end segments slightly above the bottom of plot
               color = reference_color, # set color for site marker segments
               linewidth = 0.5) + # set segment line thickness
  # add site labels with repelling so labels do not overlap
  geom_label_repel(data = label_df, # use label_df data
                   aes(x = median_do, # map label positions and label text
                       y = 0.012, # place labels near the bottom of plot
                       # wrap long site names so labels fit better
                       label = str_wrap(site_full, width = 12)),
                   size = 3, # set site label text size
                   fill = "white", # make label background white
                   color = "black", # make label text black
                   label.size = 0.2, # set thickness of label box outline
                   box.padding = 0.25, # add padding around label boxes
                   point.padding = 0.15, # padding between labels + points
                   segment.color = "gray50", # set color of repel lines
                   min.segment.length = 0, # allow connectors at length 0
                   seed = 123) + # make label placement reproducible
  # add quartile labels in boxes at the top of the plot
  geom_label(data = quartile_df, # uses quartile_df data
             aes(x = x_label, # map quartile label positions + label text
                 y = y, # place labels near the top of the plot
                 label = label), # use quartile labels created in quartile_df
             fill = "white", # make label background white 
             color = reference_color, # set color for quartile text
             label.size = 0.2, # set thickness of label box outline
             size = 3.2, # set quartile label text size
             vjust = 1) + # align labels downward from their y-position
  # add axis labels and title
  labs(x = "Median Dissolved Oxygen (mg/L)", # x-axis label
       y = "Density", # y-axis label
       # add a descriptive plot title
       title = "Distribution of Median DO Across Sites") +
  # add extra x-axis space so edge labels are not cut off
  scale_x_continuous(expand = expansion(mult = c(0.02, 0.08))) +
  # remove extra space at bottom of y-axis and set y-axis limits
  scale_y_continuous(expand = c(0, 0),
                     limits = c(0, 0.15)) + # set y-axis limits from 0 to 0.15
  # use consistent theme
  theme_ncos +
  # remove legend 
  theme(legend.position = "none")
```

After establishing this DO gradient, we examined whether total invertebrate abundance was related to continuous dissolved oxygen values. Total abundance, summed across the five focus taxa and log-transformed, showed a weak-to-moderate negative relationship with median dissolved oxygen (Figure 2).

## Figure 2. Relationship between DO and Total Invertebrate Abundance

```{r}
#| label: figure-2

# fit linear model for relationship between log abundance and median DO
total_abundance_lm <- lm(log_total_abundance ~ med_do, 
                         # use total_abundance as df
                         data = total_abundance)

# create new data frame total_abundance_predictions for predicted values
total_abundance_predictions <- data.frame( 
  # create a sequence of DO values from the minimum to maximum observed DO
  med_do = seq(min(total_abundance$med_do, na.rm = TRUE), # max value is max DO
               max(total_abundance$med_do, na.rm = TRUE), # min value is min DO
               # creates 100 evenly spaced values for smooth predictions
               length.out = 100))

# generate predicted values and confidence intervals
total_abundance_predictions <- total_abundance_predictions |>
  bind_cols( # attach model predictions + confidence interval values to df
    # predict log total abundance using the fitted linear model
    predict(total_abundance_lm, 
            # use the total_abundance_predictions df as data for prediction
            newdata = total_abundance_predictions,
            # create confidence intervals around the fitted values
            interval = "confidence") |>
      # convert prediction output into a data frame
      as.data.frame())

# creat plot from total_abundance df
ggplot(total_abundance,
       aes(x = med_do, # x-axis is total_abundance
           y = log_total_abundance)) + # y-axis is log_total_abundance
  # layer 1: create points  
  geom_point(color = raw_point_color, # color points
             size = 3, # size of points is 3
             alpha = 0.7) + # opacity of points is 0.7
  # layer 2: creates ribbon confidence interval around model predictions
  geom_ribbon(data = total_abundance_predictions, 
              aes(x = med_do, # x is median DO
                  ymin = lwr, # min on y is lower confidence interval
                  ymax = upr), # max on y is lower confidence interval
              inherit.aes = FALSE, # does not inherit ggplot aesthetics
              fill = ci_fill, # colors ribbon
              alpha = 0.35) + # opacity of points is 0.35
  # layer 3: creates predicted line from linear model
  geom_line(data = total_abundance_predictions, 
            aes(x = med_do, # x is median DO
                y = fit), # y is fitted model values
            inherit.aes = FALSE,  # does not inherit ggplot aesthetics
            color = model_color, # colors prediction line
            linewidth = 1) + # sets line width
  labs(
    x = "Dissolved oxygen (mg/L)", # labels x-axis
    y = "Log total invertebrate abundance", # labels y-axis
    # creates descriptive title
    title = "Relationship between DO and Total Invertebrate Abundance") +
  # use consistent theme
  theme_ncos
```

A Pearson correlation test showed a statistically significant negative linear relationship between median DO and log-transformed total invertebrate abundance, r = -0.375, t(32) = -2.29, p = 0.0289. The 95% confidence interval for the correlation ranged from -0.633 to -0.042, which does not include zero, further supporting a negative relationship. This means that as dissolved oxygen increased, total invertebrate abundance tended to decrease. However, because the correlation was not close to -1, the relationship was not strong; dissolved oxygen explained some, but not all, of the variation in total abundance.

```{r}
#| label: pearson-correlation-test
# run a Pearson correlation test between log total abundance + median DO
cor.test(
  total_abundance$log_total_abundance, # log total abundance is first variable
  total_abundance$med_do, # use median DO as the second variable
         method = "pearson") # specify pearson method
# r = -0.3749894, p = 0.02887
# there is a significant negative correlation between abundance and continous DO
```

We then examined whether this relationship was consistent across individual taxa. Taxon-specific correlations showed that most focus taxa had weak negative relationships with dissolved oxygen, but the strength of the relationship differed among groups (Figure 3). Cladocera and Copepod had the strongest negative correlations with DO, while Ostracod and Hemiptera Corixidae Boatman had weaker negative correlations. Nematode showed almost no relationship with dissolved oxygen. This suggests that the community-level relationship between DO and abundance was not driven equally by all taxa, but instead reflected different taxon-specific patterns.

## Figure 3: Correlation between DO & Invertebrate Abundance

```{r}
#| label: figure-3
# create wide dataset (one column per taxon)
taxa_wide <- DO_taxa |>
  # keep only the columns needed 
  select(date, site, taxon, log_ave_lit, med_do) |>
  # convert the dataset from long format to wide format
  pivot_wider(names_from = taxon, # create one new column for each taxon
              # fill each taxon column with log-transformed abundance values
              values_from = log_ave_lit)

# calculate correlation of each taxon with dissolved oxygen
# calculate Spearman correlation of each taxon with dissolved oxygen
cor_do <- taxa_wide |>
  # remove date and site columns so only DO and taxon abundance columns remain
  select(-date, -site) |>
  # calculate correlations for each taxon column
  summarize(across(-med_do, ~ cor(.x,
                                  # correlate each column w/ median DO
                                  med_do,
                                  # specify spearman method
                                  method = "spearman",
                                  # only use rows w/ complete observations 
                                  use = "complete.obs"))) |>
  # convert the correlation results from wide format to long format
  pivot_longer(cols = everything(),
               # store taxon names in a column called taxon
               names_to = "taxon", 
               # store correlation values in a column called correlation
               values_to = "correlation")

ggplot(cor_do, # create a new plot from cor_do
       aes(x = taxon, # x-axis is taxon
           y = "Dissolved oxygen", # titles y-axis
           fill = correlation)) + # fill color by correlation
  # add tiles where fill represents the strength + direction of correlation
  geom_tile(color = "grey70") +
  # add circles sized by the absolute value of the correlation
  geom_point(aes(size = abs(correlation)), shape = 21, fill = "white") +
  # add labels above circles
  geom_text(aes(label = round(correlation, 2)),
            # move the text labels slightly above the circles
            nudge_y = 0.13, size = 4.5) +
  # create a color scale for negative, neutral, and positive correlations
  scale_fill_gradient2(
    # use red for negative correlations
    low = do_colors["low"],
    # use white for correlations near zero
    mid = do_colors["moderate"],
    # use blue for positive correlations
    high = do_colors["high"],
    # sets midpoint to 0
    midpoint = 0,
    # set the color scale limits from -1 to 1 
    limits = c(-1, 1),
    # labels legend 'Spearman r'
    name = "Spearman r") +
  # set the range of circle sizes + remove the size legend
  scale_size(range = c(2, 10), guide = "none") +
  # wrap long taxon labels
  scale_x_discrete(labels = \(x) stringr::str_wrap(x, width = 12)) +
  labs(x = "Taxon", # labels x-axis
       y = NULL, # remove the y-axis label
       # creates descriptive title
       title = "Correlation between DO & Invertebrate Abundance") +
  # use consistent theme
  theme_ncos +
  theme(axis.text.y = element_text(size = 10), # change y axis text size
        plot.title = element_text(hjust = 0.5), # adjust title alignment
        panel.grid = element_blank()) # remove grid
```

To compare abundance across broader oxygen conditions, we grouped observations into low, medium, and high dissolved oxygen categories. Total log-transformed invertebrate abundance appeared lower in the High DO category than in the Low DO and Medium DO categories (Figure 4). However, there was substantial overlap among the categories, especially in the Medium DO group. 

## Figure 4: Total Invertebrate Abundance Across DO Levels

```{r}
#| label: figure-4
ggplot(total_abundance, # uses total_abundance df
       aes(x = do_category, # x-axis is do_category
           y = log_total_abundance, # y-axis is log_total_abundance
           fill = do_category)) + # fill color by do_category
  # layer 1: add boxplots to summarize abundance distribution w/in each DO level 
  geom_boxplot(alpha = 0.7, # transparency is 0.7
               outlier.color = raw_point_color, # color outlier points 
               outlier.size = 2) + # outlier size is 2
  # layer 2: adds jittered points of raw data
  geom_jitter(width = 0.15, # sets point width
              alpha = 0.5, # sets point opacity
              size = 2, # sets point size
              color = raw_point_color) + # sets point color
  labs(
    x = "DO level", # labels x-axis
    y = "Log total invertebrate abundance", # labels y-axis
    # creates descriptive title
    title = "Total Invertebrate Abundance Across DO Levels") +
  # manually assign fill colors to each DO category
  scale_fill_manual(values = c(
    # use red for low DO bin
    "Low DO" = 'red3',
    # use white for low mid bin
    "Medium DO" = 'grey',
    # use blue for low DO bin
    "High DO" = 'blue3')) +
  # uses consistent theme
  theme_ncos +
  # removes redundant legend
  theme(legend.position = "none")
```

A Kruskal-Wallis test showed that total abundance did not differ significantly across the three DO categories, X² = 2.00, df = 2, p = 0.368. This indicates that, although the visualization suggests lower abundance under high DO conditions, there was not enough statistical evidence to conclude that total abundance differed significantly among the DO categories.

```{r}
#| label: Kruskal-Wallis-test
# run a Kruskal-Wallis test comparing log total abundance across DO categories
kruskal.test(log_total_abundance ~ do_category,
             # use the total_abundance dataset
             data = total_abundance)
```

Taxon-specific abundance patterns across DO categories showed more detail than the community-level comparison. Copepods and Ostracods had higher and more variable abundances in the Moderate DO category, while Cladocera showed some high-abundance observations in the Low and Moderate DO categories (Figure 5). Nematodes remained close to zero across all dissolved oxygen levels, and Hemiptera Corixidae Boatman generally had low abundance, with only a few higher observations in Moderate DO. These patterns suggest that dissolved oxygen may be related to abundance differently depending on the taxon, rather than producing one consistent response across the whole invertebrate community.

## Figure 5. Taxon-Specific Invertebrate Abundance Across DO Levels

```{r}
#| label: figure-5
#| fig-width: 8
#| fig-height: 5

# base layer: ggplot
ins_bins_plot <- ggplot(data = aq_ins_clean_bins, # uses aq_ins_clean_bins df
                        mapping = aes(x = bins, # x-axis: DO bins
                                      y = log_ave_lit, # y-axis: log_ave_lit
                                      color = taxon)) + # color by taxon
  # first layer: points show individual abundance measurements 
  geom_jitter(width = 0.12, # sets jitter width
              height = 0, # sets jitter height
              shape = 16, # sets jitter point shape
              size = 2, # sets jitter point size
              alpha = 0.7) + # sets jitter point opacity
  # calculates median for each bin for each taxon
  stat_summary(fun = median, # uses median function
               geom = "point", # uses point geometry
               color = "red3", # colors median points red
               size = 3) + # sets median point size
  # creates separate panels for each taxon 
  facet_wrap(~ taxon,
             # wrap long strip labels so they fit inside each panel
             labeller = labeller(taxon = label_wrap_gen(width = 20))) +
  # uses consistent colors for points
  scale_color_manual(values = taxon_colors) +
  # relabel the x-axis categories to improve readability
  scale_x_discrete(labels = c("low" = "Low", 
                              "moderate" = "Moderate", 
                              "high" = "High")) +
  labs(x = "DO levels", # relabel x-axis
       y = "log + 1 transformed average abundance/L", # relabel y-axis
       # creates descriptive title
       title = "Taxon-Specific Invertebrate Abundance Across DO Levels") +
  # uses consistent theme
  theme_ncos +
  theme(legend.position = "none", # remove redundant legend
        # make facet strip text slightly smaller so long names fit better
        strip.text = element_text(size = 10),
        # make x-axis text slightly smaller
        axis.text.x = element_text(size = 9),
        # make y-axis text slightly smaller
        axis.text.y = element_text(size = 9))

# shows plot
ins_bins_plot
```

Finally, we examined whether the relative composition of the invertebrate community changed across dissolved oxygen categories. Community composition shifted clearly among Low, Medium, and High DO groups (Figure 6). Low DO sites had a more mixed community dominated mostly by Ostracods and Cladocera. Moderate DO sites were dominated by Copepods and Ostracods, with smaller contributions from Cladocera and Hemiptera Corixidae Boatman. High DO sites were dominated primarily by Copepods, with Cladocera making up a smaller portion of the community and Ostracods contributing very little. Hemiptera Corixidae Boatman and Nematodes made up small proportions of the community across all DO categories.

## Figure 6. Community Composition Across DO Levels

```{r}
#| label: figure-6
# creates obejct composition_plot using DO_taxa df
composition_plot <- DO_taxa |> 
  mutate(
    # classify each observation into low, medium, or high dissolved oxygen
    do_category = case_when(
      # assign Low DO to values at or below the 25th percentile of median DO
      med_do <= quantile(med_do, 0.25, na.rm = TRUE) ~ "Low DO",
      # assign Medium DO to values between the 25th percentile - 75th percentile
      med_do <= quantile(med_do, 0.75, na.rm = TRUE) ~ "Medium DO",
      # assign High DO to all remaining values above the 75th percentile
      TRUE ~ "High DO"),
    # convert do_category into a factor to control plotting order
    do_category = factor(do_category,
                         # set order of DO categories on the x-axis
                         levels = c("Low DO", "Medium DO", "High DO"))) |>
  # groups by do_category and taxon
  group_by(do_category, taxon) |>
  # calculate total abundance for each taxon within each DO category
  summarize(
    # sum average abundance per liter + ignore missing values
    total_abundance = sum(ave_lit, na.rm = TRUE),
    # ungroup
    .groups = "drop") |>

# create plot from composition_plot df
ggplot(aes(x = do_category, # x-axis is do_category
           y = total_abundance, # y-axis is total_abundance
           fill = taxon)) + # fill color by taxon
  # create stacked bars scaled to proportions within each DO category
  geom_col(position = "fill") + 
  # use consistent colors
  scale_fill_manual(values = taxon_colors) +
  # format the y-axis as percentages
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    x = "Dissolved oxygen category", # relabels x-axis
    y = "Relative community composition", # relabels y-axis
    fill = "Taxon", # labels legend
    title = "Community Composition Across DO Levels") +
  # use consistent theme
  theme_ncos

composition_plot
```

A Pearson’s chi-squared test showed that taxon composition differed significantly across DO bins, X² = 2859.2, df = 8, p < 2.2e-16. This provides strong evidence that the relative abundances of the focus taxa were not evenly distributed across low, medium, and high DO conditions. In other words, dissolved oxygen categories were strongly associated with changes in community composition, even though total abundance did not differ significantly across DO categories.

```{r}
#| label: Pearson's-chi-square
# create abundance table for chi-square test
chi_square_table <- aq_ins_clean_bins |>
  # keep only rows with the focus taxa
  filter(taxon %in% focus_taxa) |>
  # group data by DO bin + taxon
  group_by(bins, taxon) |>
  summarise(
    # sum average abundance per liter + ignore missing values
    total_abundance = sum(ave_lit, na.rm = TRUE),
    # ungroup
    .groups = "drop") |>
  # convert the table from long format to wide format
  pivot_wider(
    # create one column for each taxon
    names_from = taxon,
    # fill taxon columns with total abundance values
    values_from = total_abundance,
    # fill missing values with 0
    values_fill = 0) |>
  # convert the bins column into row names
  column_to_rownames("bins") |>
  # convert the table into a matrix
  as.matrix()

# show table
chi_square_table

# chi-square test for differences in taxon composition across DO bins
chi_square_test <- chisq.test(chi_square_table)

# show chi-square result
chi_square_test
```

Overall, the results suggest that the relationship between dissolved oxygen and invertebrate abundance at NCOS was weak-to-moderate and generally negative. Total invertebrate abundance decreased as DO increased when DO was analyzed as a continuous variable, but abundance differences across broad DO categories were not statistically significant. The strongest evidence of a DO-related pattern was seen in community composition, where the relative abundance of different taxa shifted strongly across oxygen categories. This indicates that dissolved oxygen may be more strongly related to which taxa dominate the invertebrate community than to total abundance alone.


# Discussion

# References