---
title: "Timeline Check-In"
format: pdf
execute: 
  message: false
  warning: false
---
# Set Up
## Packages

```{r}
#| label: packages

# general use and cleaning
library(tidyverse)
library(here)
library(janitor)
library(snakecase)

# wrap axis labels
library(scales)

# read .xlsx files
library(readxl)

# visualize missing data
library(naniar)

# arrange plots
library(patchwork)

library(dplyr)     # data manipulation
library(broom)     # tidy statistical output
library(corrplot)
library(ggrepel)

```



## Data

```{r}
#| label: data

# taxonomic information
taxon_list <- read_csv(here("data", "taxon_list.csv"))

# aquatic invertebrates
aq_ins <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                    sheet = "Aquatic Insects")


# read in site information
sites <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                   sheet = "sites")

# read in water quality information
water_quality <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                           sheet = "Water Quality",
                           na = c("", "n/a", "N/A", "over", "173+")) 
```

## Cleaning
### `sites` object
```{r}
#| label: ncos-site-object

# creating new object from sites
ncos_sites <- sites |> 
  # filter to only include NCOS sites sampled four times a year
  filter(sector == "NCOS" & sampling_frequency == "Quarterly")
```
### Cleaning `taxon_list`
```{r}
#| label: cleaning-taxon-list

# creating new clean object from taxon_list
taxon_list_clean <- taxon_list |> 
  # converting taxon name to snake case (no spaces or capital letters,
  # only underscores)
  mutate(verbatim_name = to_any_case(verbatim_name, case = "snake"))
```
### Cleaning `water_quality`
```{r}
#| label: water-quality-cleaning

# creating new clean object from water quality
water_quality_clean <- water_quality |> 
  # clean column names
  clean_names() |> 
  # filtering join: only include sites in ncos_sites
  semi_join(ncos_sites, by = "site") |> 
  # create new column to join with later data frames
  unite("date_site", date, site, remove = TRUE) |> 
  # group by date_site column
  group_by(date_site) |>
  # calculate median pH, salinity, DO
  summarize(med_ph = median(dissolved_oxygen_mg_l, na.rm = TRUE),
            med_sal = median(salinity_ppt, na.rm = TRUE),
            med_do = median(dissolved_oxygen_mg_l, na.rm = TRUE)) |> 
  # ungroup data frame
  ungroup() |> 
  # filter out salinity outliers
  filter(date_site != "2022-08-12_NVBR") 
```

### Cleaning `aq_ins`
```{r}
#| label: cleaning-aq-ins

# creating new clean object from aquatic inverts
aq_ins_clean <- aq_ins |> 
  # clean column names
  clean_names() |> 
  # filtering join: only include sites occurring in ncos_sites
  semi_join(ncos_sites, by = c("site")) |> 
  # renaming columns
  rename(date = date_on_vial) |> 
  # select columns of interest
  select(date, sample_type, site,
         ostracod,
         copepod,
         hemiptera_corixidae_boatman,
         cladocera,
         nematode,
         diptera_ceratopogonidae, 
         annelida_oligochaete,
         diptera_chironomid,
         annelida_polychaete) |> 
  # filter to only include "filtered beaker" samples
  filter(sample_type %in% c("FB 250", "FB250")) |> 
  # convert the data frame to long format
  pivot_longer(cols = ostracod:annelida_polychaete,
               names_to = "taxon",
               values_to = "abundance") |> 
  # group by date, site, taxon
  group_by(date, site, taxon) |> 
  # calculate average abundance per liter (sum abundance divided by 7.5 liters)
  summarize(ave_lit = sum(abundance, na.rm = TRUE)/7.5) |> 
  # ungroup data frame
  ungroup() |> 
  # create a new column called `date_site` to join with water quality
  unite("date_site", date, site, remove = FALSE) |> 
  # join with water quality
  left_join(water_quality_clean, by = c("date_site")) |> 
  # join with taxon list
  left_join(taxon_list_clean, by = c("taxon" = "verbatim_name")) |> 
  # add full names of sites
  mutate(site_full = case_when(
    site == "NVBR" ~ "North of Venoco Bridge",
    site == "NEC" ~ "South of Venoco Bridge",
    site == "NMC" ~ "Main Channel",
    site == "NPB" ~ "South of Phelps Bridge",
    site == "NPB1" ~ "North of Phelps Bridge",
    site == "NPB2" ~ "Phelps Road",
    site == "NWP" ~  "West Pond",
    site == "NDC" ~ "Devereux Creek"
  )) |> 
  # setting factor levels for site
  mutate(site_full = fct_reorder(site_full, med_do, .fun = "median", .na_rm = TRUE),
         site_full = fct_rev(site_full))
```
### Focus Taxa (`focus_taxa`)
```{r}
#| label: focus_taxa
# taxa to include in all figures
focus_taxa <- c("Ostracod", "Copepod", "Cladocera", 
                "Hemiptera Corixidae Boatman", "Nematode")
```

### Creating DO bins (creating `do_category` column)
```{r}
aq_ins_clean_bins <- aq_ins_clean |> 
  mutate(bins = case_when(site_full == "Phelps Road" ~ "low",
                             site_full == "North of Phelps Bridge" ~ "low", 
                             site_full == "South of Venoco Bridge" ~ "moderate",
                             site_full == "South of Phelps Bridge" ~ "moderate",
                             site_full =="Devereux Creek" ~ "moderate",
                             site_full =="North of Venoco Bridge" ~ "high",
                             site_full == "Main Channel" ~ "high")) |> 
  mutate(taxon = str_replace_all(taxon, "_", " "), taxon = str_to_title(taxon)) |>
  filter(taxon %in% focus_taxa) |>
  # create column that calculates log-transformed abundance
  mutate(log_ave_lit = log(ave_lit + 1)) |> 
  mutate(bins = factor(bins, levels = c("low", "moderate", "high")))

# create object DO_taxa from aq_ins_clean
DO_taxa <- aq_ins_clean |> 
  # remove rows that are missing median salinity values
  filter(!is.na(med_do)) |> 
  # create column that calculates log-transformed abundance
  mutate(log_ave_lit = log(ave_lit + 1)) |>
  # mutates to title-case taxon names
  mutate(taxon = str_replace_all(taxon, "_", " "), taxon = str_to_title(taxon)) |>
  # only include focus taxa
  filter(taxon %in% focus_taxa) |> 
  # order taxon levels by maximum average abundance per liter
  mutate(taxon = fct_reorder(taxon, ave_lit, .fun = max, .desc = TRUE)) |> 
  # keep only the columns needed for the figure
  select(date, site, site_full, taxon, ave_lit, log_ave_lit, med_do)
total_abundance <- DO_taxa |>
  filter(taxon %in% focus_taxa) |>
  group_by(date, site, med_do) |>
  summarize(
    total_abundance = sum(ave_lit, na.rm = TRUE),
    log_total_abundance = log10(total_abundance + 1),
    .groups = "drop"
  )
total_abundance <- total_abundance |>
  mutate(
    do_category = case_when(
      med_do <= quantile(med_do, 0.25, na.rm = TRUE) ~ "Low DO",
      med_do <= quantile(med_do, 0.75, na.rm = TRUE) ~ "Medium DO",
      TRUE ~ "High DO"
    )
  )
total_abundance$do_category <- factor(
  total_abundance$do_category,
  levels = c("Low DO", "Medium DO", "High DO")
)
```

# Ideas for analysis and background

- should include some citations (not all)
- should include ideas for analyses to answer questions and some preliminary code to run those analyses

## Citations:

NEED TO ADD

## Spearman Correlation Analysis for Total Abundance and DO levels
```{r}
cor.test(
  total_abundance$log_total_abundance,
  total_abundance$med_do,
  method = "spearman",
  exact = FALSE)
```
Total log-transformed invertebrate abundance had a **weak negative** relationship with dissolved oxygen, but this relationship was not statistically significant using Spearman correlation.

## Correlation analysis for Each Focus Taxa and DO levels
```{r}
#| fig-width: 10
#| fig-height: 8
# create wide dataset (one column per taxon)
taxa_wide <- DO_taxa |>
  select(date, site, taxon, log_ave_lit, med_do) |>
  pivot_wider(
    names_from = taxon,
    values_from = log_ave_lit
  )

# calculate correlation of each taxon with dissolved oxygen
# calculate Spearman correlation of each taxon with dissolved oxygen
cor_do <- taxa_wide |>
  select(-date, -site) |>
  summarize(across(-med_do, ~ cor(.x,
                                  med_do,
                                  method = "spearman",
                                  use = "complete.obs"))) |>
  pivot_longer(cols = everything(),
               names_to = "taxon",
               values_to = "correlation")

ggplot(cor_do, aes(x = taxon, y = "Dissolved oxygen", fill = correlation)) +
  geom_tile(color = "grey70") +
  geom_point(aes(size = abs(correlation)), shape = 21, fill = "white") +
  
  # move labels ABOVE circles
  geom_text(aes(label = round(correlation, 2)),
            nudge_y = 0.15,  # <-- this moves text up
            size = 4) +
  
  scale_fill_gradient2(
    low = "red",
    mid = "white",
    high = "blue",
    midpoint = 0,
    limits = c(-1, 1),
    name = "r"
  ) +
  scale_size(range = c(2, 10), guide = "none") +
  labs(
    x = "Taxon",
    y = NULL,
    title = "Spearman correlation between dissolved oxygen and invertebrate abundance"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```
Most focus taxa showed **weak negative** correlations with dissolved oxygen, with Cladocera and Copepod showing the strongest negative relationships and Nematode showing almost no relationship.

## Pearson correlation for continuous DO and total community abundance
```{r}
# Pearson correlation for continuous DO and total community abundance
cor.test(total_abundance$log_total_abundance,
         total_abundance$med_do,
         method = "pearson") 
# r = -0.3749894, p = 0.02887
# there is a significant negative correlation between abundance and continous DO
```
There was a **statistically significant negative linear relationship** between median dissolved oxygen and log-transformed total invertebrate abundance, suggesting that abundance decreased as DO increased.

## Kruskal-Wallis test comparing abundance across DO categories
```{r}
kruskal.test(log_total_abundance ~ do_category,
             data = total_abundance)
# Kruskal-Wallis chi-squared = 3.2212, df = 2, p-value = 0.1998
# there is not a significant difference between abundance and DO categories
```
Total log-transformed invertebrate abundance **did not differ significantly** across low, moderate, and high DO categories.

## Kruskal-Wallis test between abundance and bins for Copepods
```{r}
kruskal.test(log_ave_lit ~ bins,
             data = aq_ins_clean_bins,
             subset = taxon == "Copepod") 
# Kruskal-Wallis chi-squared = 4.405, df = 2, p-value = 0.1105
# no significant difference
```
Copepod abundance **did not differ significantly across DO bins**, suggesting DO category was not a strong predictor of Copepod abundance.

## Kruskal-Wallis test between abundance and bins for Ostracods
```{r}
kruskal.test(log_ave_lit ~ bins,
             data = aq_ins_clean_bins,
             subset = taxon == "Ostracod") 
# Kruskal-Wallis chi-squared = 6.7235, df = 2, p-value = 0.03467
# significant difference of abundance btwn DO level bins
```                                    
Ostracod abundance **differed significantly across DO bins**, indicating that Ostracods may respond differently to low, moderate, and high DO conditions.

## Pairwise test between abundance and individual bins for Ostracods
```{r}
bins_ostracods <- aq_ins_clean_bins |> 
  filter(taxon == "Ostracod")
pairwise.wilcox.test(
  bins_ostracods$log_ave_lit,
  bins_ostracods$bins,
  p.adjust.method = "bonferroni") 
# p = 0.051 for comparison btwn moderate and high DO level bins
```
After Bonferroni correction, the moderate vs. high DO comparison was close to significant, suggesting a possible difference in Ostracod abundance between these bins but not strong enough to conclude significance at $\alpha$ = 0.05.

## Chi-Square
```{r}
# create abundance table for chi-square test
chi_square_table <- aq_ins_clean_bins |>
  filter(taxon %in% focus_taxa) |>
  group_by(bins, taxon) |>
  summarise(
    total_abundance = sum(ave_lit, na.rm = TRUE),
    .groups = "drop"
  ) |>
  pivot_wider(
    names_from = taxon,
    values_from = total_abundance,
    values_fill = 0
  ) |>
  column_to_rownames("bins") |>
  as.matrix()

view(chi_square_table)

# show table used in chi-square test
chi_square_table

# chi-square test for differences in taxon composition across DO bins
chi_square_test <- chisq.test(chi_square_table)

# show chi-square result
chi_square_test
```
Taxon composition **differed significantly across DO bins**, suggesting that the invertebrate community structure changes depending on dissolved oxygen category.

# Visualizations directly relevant to answering questions
## Directly Relevant Figure 1: DO levels vs invert. abundance for each focus taxa
```{r}
#| label: wrangle_data

DO_taxa_medians <- DO_taxa |>
  # groups data by taxon so each species gets one median point
  group_by(taxon) |>
  # calculates median DO and median log abundance for each taxon
  summarise(
    med_do_point = median(med_do, na.rm = TRUE),
    med_log_ave_lit = median(log_ave_lit, na.rm = TRUE),
    .groups = "drop"
)
```
```{r}
#| label: DO_taxa_plot
#| fig-width: 10
#| fig-height: 8

DO_taxa_plot <- ggplot(data = DO_taxa, # assigns plot to object do_taxa_plot 
                        # uses DO_taxa data frame
                        mapping = aes(x = med_do, # x-axis: med_do
                                      y = log_ave_lit, # y-axis: log_ave_lit
                                      color = taxon)) + # color by taxon
  # first layer: points show individual abundance measurements 
  geom_point(shape = 16, # shape is circle
             size = 2, # size is 2
             alpha = 0.7) + # opacity is 0.7
  
  # second layer: larger point shows the median location for each taxon
  geom_point(data = DO_taxa_medians,
             mapping = aes(x = med_do_point,
                           y = med_log_ave_lit),
             shape = 18,
             size = 5) +
  
  # separate panels for each taxon with free x- and y-axis scales
  facet_wrap(~ taxon, scales = "free") +
  # use non-default colors for points
  scale_color_viridis_d() +
  labs(x = "Median DO (mg/L)", # relabel x-axis
       y = "log + 1 transformed average abundance/L", # relabel y-axis
       # creates descriptive title
       title = "Invertebrate abundance as a function of DO") +
  
  # uses non-default theme
  theme_minimal() +
  # remove redundant legend
  theme(legend.position = "none")
DO_taxa_plot
```
## Directly Relevant Figure 2: DO levels vs total invert. abundance
```{r}
#| fig-width: 10
#| fig-height: 8
# Summed abundance across focus taxa vs DO
focus_taxa <- c("Cladocera", "Copepod", 
                "Hemiptera Corixidae Boatman", 
                "Nematode", "Ostracod")

total_abundance <- DO_taxa |>
  filter(taxon %in% focus_taxa) |>
  group_by(date, site, med_do) |>
  summarize(
    total_abundance = sum(ave_lit, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(log_total_abundance = log10(total_abundance + 1))

# fit linear model
total_abundance_lm <- lm(log_total_abundance ~ med_do,
                         data = total_abundance)

# create new data frame for predicted values
total_abundance_predictions <- data.frame(
  med_do = seq(min(total_abundance$med_do, na.rm = TRUE),
               max(total_abundance$med_do, na.rm = TRUE),
               length.out = 100)
)

# generate predicted values and confidence intervals
total_abundance_predictions <- total_abundance_predictions |>
  bind_cols(
    predict(total_abundance_lm,
            newdata = total_abundance_predictions,
            interval = "confidence") |>
      as.data.frame()
  )

# plot raw data and model predictions
ggplot(total_abundance,
       aes(x = med_do,
           y = log_total_abundance)) +
  
  # raw data points
  geom_point(color = "magenta3",
             size = 3,
             alpha = 0.7) +
  
  # confidence interval around model predictions
  geom_ribbon(data = total_abundance_predictions,
              aes(x = med_do,
                  ymin = lwr,
                  ymax = upr),
              inherit.aes = FALSE,
              fill = "cyan2",
              alpha = 0.35) +
  
  # predicted line from linear model
  geom_line(data = total_abundance_predictions,
            aes(x = med_do,
                y = fit),
            inherit.aes = FALSE,
            color = "purple4",
            linewidth = 1) +
  
  labs(
    x = "Dissolved oxygen (mg/L)",
    y = "Log total invertebrate abundance",
    title = "Relationship between dissolved oxygen and total invertebrate abundance"
  ) +
  
  theme_classic()
```
```{r}
broom::tidy(total_abundance_lm)
```
## Directly Relevant Figure 3: Abundance vs DO with binned levels
```{r}
total_abundance <- total_abundance |>
  mutate(
    do_category = case_when(
      med_do <= quantile(med_do, 0.25, na.rm = TRUE) ~ "Low DO",
      med_do <= quantile(med_do, 0.75, na.rm = TRUE) ~ "Medium DO",
      TRUE ~ "High DO"
    )
  )
total_abundance$do_category <- factor(
  total_abundance$do_category,
  levels = c("Low DO", "Medium DO", "High DO")
)

ggplot(total_abundance,
       aes(x = do_category,
           y = log_total_abundance,
           fill = do_category)) +
  
  geom_boxplot(alpha = 0.7,
               outlier.color = "black",
               outlier.size = 2) +
  
  # optional: add jittered points to show raw data
  geom_jitter(width = 0.15,
              alpha = 0.5,
              size = 2,
              color = "black") +
  
  labs(
    x = "Dissolved oxygen level",
    y = "Log total invertebrate abundance",
    title = "Invertebrate abundance across dissolved oxygen levels"
  ) +
  
  scale_fill_manual(values = c(
    "Low DO" = "red3",
    "Medium DO" = "grey60",
    "High DO" = "blue3"
  )) +
  
  theme_classic() +
  theme(legend.position = "none")
```

# Other exploratory visualizations
## Exploratory Figure 1: Variance of DO across sights
```{r}
#| fig-width: 10
#| fig-height: 8
# create object aq_ins_clean_fig2 from aq_ins_clean
aq_ins_clean_fig2 <-  aq_ins_clean |>
  # convert taxon to title case
  mutate(taxon = str_replace_all(taxon, "_", " "),
         taxon = str_to_title(taxon)) |> 
  # filter to include focus taxa
  filter(taxon %in% focus_taxa)

# base layer
DO <- ggplot(data = aq_ins_clean_fig2,
       # x-axis: site
       # y-axis: mean salinity
       mapping = aes(x = site_full,
                     y = med_do)) +
  # first layer: boxplot
  geom_boxplot(width = 0.35,
               outlier.shape = NA) +
  # second layer: salinity measurements at each survey
  geom_jitter(height = 0,
              width = 0.12,
              shape = 21,
              alpha = 0.6) +
  # wrapping x-axis text
  scale_x_discrete(labels = label_wrap(width = 15)) +
  # second layer: points to represent means
  stat_summary(fun = median,
               color = "red",
               size = 1) +
  # relabelling x- and y-axes
  labs(x = "Site",
       y = "Median DO (mh/L)",
       title = "Site DO levels") +
  # different theme
  theme_bw()
# show plot
DO
```
## Exploratory Figure 2: Composition plot across DO bins
```{r}
#| fig-width: 10
#| fig-height: 8
composition_plot <- DO_taxa |>
  mutate(
    do_category = case_when(
      med_do <= quantile(med_do, 0.25, na.rm = TRUE) ~ "Low DO",
      med_do <= quantile(med_do, 0.75, na.rm = TRUE) ~ "Medium DO",
      TRUE ~ "High DO"
    ),
    do_category = factor(do_category,
                         levels = c("Low DO",
                                    "Medium DO",
                                    "High DO"))
  ) |>
  group_by(do_category, taxon) |>
  summarize(
    total_abundance = sum(ave_lit, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(x = do_category,
             y = total_abundance,
             fill = taxon)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    x = "Dissolved oxygen category",
    y = "Relative community composition",
    fill = "Taxon",
    title = "Invertebrate community composition shifts across dissolved oxygen levels"
  ) +
  theme_classic()

# display plot
composition_plot
```
## Exploratory Figure 3: Time-series plot of total abundance (all species)
```{r}
# species/taxa to include in the total abundance figure
species_interest <- c(
  "Ostracod",
  "Copepod",
  "Cladocera",
  "Hemiptera Corixidae Boatman",
  "Nematode"
)

# create total abundance over time
total_abundance_time <- DO_taxa |>
  # keeps only the taxa of interest
  filter(taxon %in% species_interest) |>
  # makes sure date is treated as a date
  mutate(date = as.Date(date)) |>
  # groups by date only, so abundance is summed across sites and taxa
  group_by(date) |>
  # sums average abundance per liter across all taxa of interest
  summarise(
    total_abundance = sum(ave_lit, na.rm = TRUE),
    .groups = "drop"
  )

# plot total abundance over time
ggplot(total_abundance_time, aes(x = date, y = total_abundance)) +
  # adds a line connecting total abundance values over time
  geom_line(linewidth = 1) +
  # adds points for each sampling date
  geom_point(size = 2) +
  # formats y-axis labels with commas
  scale_y_continuous(labels = scales::comma) +
  # adds clear labels
  labs(
    title = "Total Abundance of Species of Interest Over Time",
    x = "Sampling Date",
    y = "Total Abundance"
  ) +
  # uses a clean theme
  theme_minimal()
```
## Exploratory Figure 4: Time-series area plot
```{r}
# species/taxa to include in the area plot
species_interest <- c(
  "Ostracod",
  "Copepod",
  "Cladocera",
  "Hemiptera Corixidae Boatman",
  "Nematode"
)

# create abundance over time by taxon
taxon_abundance_time <- DO_taxa |>
  # keeps only the taxa of interest
  filter(taxon %in% species_interest) |>
  # makes sure date is treated as a date
  mutate(date = as.Date(date)) |>
  # groups by date and taxon so each species stays separate
  group_by(date, taxon) |>
  # sums across sites, but not across taxa
  summarise(
    ave_lit = sum(ave_lit, na.rm = TRUE),
    .groups = "drop"
  )

# create time series area plot by taxon
ggplot(taxon_abundance_time, aes(x = date, y = ave_lit, fill = taxon)) +
  # creates stacked area plot with a different color for each taxon
  geom_area(alpha = 0.8) +
  # formats y-axis labels with commas
  scale_y_continuous(labels = scales::comma) +
  # adds clear labels
  labs(
    title = "Abundance of Species of Interest Over Time",
    x = "Sampling Date",
    y = "Average Abundance per Liter",
    fill = "Taxon"
  ) +
  # uses a clean theme
  theme_minimal()
```
## Exploratory Figure 5: Distribution of Median Dissolved Oxygen Aross Sites
```{r}
#| label: do-vs-sites-bins
#| fig-width: 10
#| fig-height: 8

# create new object to store medians of median do levels
median_fig2 <- aq_ins_clean_fig2 |> 
  group_by(site_full) |> 
  summarize(median_do = median(med_do, na.rm = TRUE))

# assign quartiles for medians for each site
median_fig2$quartile <- cut(
  median_fig2$median_do, 
  breaks = quantile(median_fig2$median_do, probs = seq(0, 1, 0.25)), 
  labels = c("Q1", "Q2", "Q3", "Q4"), 
  include.lowest = TRUE,
  right = FALSE)

# boundaries of quartiles
q <- quantile(median_fig2$median_do, probs = seq(0, 1, 0.25))

# distribution plot of DO levels across sites
ggplot(median_fig2, aes(x = median_do)) +
  # density curve
  geom_density(fill = "lightblue",
               alpha = 0.4,
               color = "blue",
               linewidth = 1.2) +
  # vertical quartile lines
  geom_vline(xintercept = q,
             linetype = "dashed",
             color = "gray40") +
  # site labels
  geom_text_repel(aes(y = 0.01,
                      label = str_wrap(site_full, width = 10)),
                  size = 3,
                  box.padding = 0.5,       # space around the text box
                  show.legend = FALSE) +
  # quartile dashed line labels
  geom_text(data = data.frame(x = q, label = names(q)), 
            aes(x = x, y = 0.14, label = paste0(label, "\n", round(x, 2))),
            color = "gray40",
            size = 3.5,
            hjust = -0.1) + # shift text slightly to the right of the line
  # remove space between bottom of plot and y-axis, set y-axis limits
  scale_y_continuous(expand = c(0,0),
                     limits = c(0, 0.15)) +
  # axes labels and title
  labs(x = "Median Dissolved Oxygen (mg/L)",
       y = "Density",
       title = "Distribution of Median Dissolved Oxygen Aross Sites") +
  # theme
  theme_bw() +
  theme(legend.position = "none")
```
These bins were created by taking the median value of the median dissolved oxygen levels at each site. Low DO is the 25th percentile, moderate DO is the 75th percentile, and high DO is the 100th percentile.

- Low DO: Phelps Road and North of Phelps Bridge
- Moderate DO: South of Venoco Bridge, South of Phelps Bridge, Devereux Creek
- High DO: North of Venoco Bridge, Main Channel

## Exploratory Figure 6: Invertebrate Abundance in DO Bins
```{r}
#| fig-width: 10
#| fig-height: 8
#| label: invert-abundance-at-bins

aq_ins_clean_bins <- aq_ins_clean |> 
  mutate(bins = case_when(site_full == "Phelps Road" ~ "low",
                             site_full == "North of Phelps Bridge" ~ "low", 
                             site_full == "South of Venoco Bridge" ~ "moderate",
                             site_full == "South of Phelps Bridge" ~ "moderate",
                             site_full =="Devereux Creek" ~ "moderate",
                             site_full =="North of Venoco Bridge" ~ "high",
                             site_full == "Main Channel" ~ "high")) |> 
  mutate(taxon = str_replace_all(taxon, "_", " "), taxon = str_to_title(taxon)) |>
  filter(taxon %in% focus_taxa) |>
  # create column that calculates log-transformed abundance
  mutate(log_ave_lit = log(ave_lit + 1)) |> 
  mutate(bins = factor(bins, levels = c("low", "moderate", "high")))
  

# base layer: ggplot
ins_bins_plot <- ggplot(data = aq_ins_clean_bins, 
                        mapping = aes(x = bins, # x-axis: DO bins
                                      y = log_ave_lit, # y-axis: log_ave_lit
                                      color = taxon)) + # color by taxon
  # first layer: points show individual abundance measurements 
  geom_jitter(width = 0.15,
            height = 0,
            shape = 16,
            size = 2,
            alpha = 0.7) +
  # calculate median for each bin for each taxon
  stat_summary(fun = median,
               geom = "point",
               color = "red",
               size = 3) +
  # separate panels for each taxon 
  facet_wrap(~ taxon) +
  # use non-default colors for points
  scale_color_viridis_d() +
  labs(x = "DO levels", # relabel x-axis
       y = "log + 1 transformed average abundance/L", # relabel y-axis
       # creates descriptive title
       title = "Invertebrate abundance vs DO levels") +
  # uses non-default theme
  theme_bw() +
  # remove redundant legend
  theme(legend.position = "none")

# shows plot
ins_bins_plot

```


# Plan for elective (if completing)
- should include some description/image of drafts
- should include plan for making progress in the next two weeks